\Large Analyzing the Supreme Court Citation Network

### 2017-09-13 Tex/paper.tex
MCMLE estimation and MH algorithm

 \section*{The Exponential Random Configuration Model}
Let $c(t)\in \{0,1\}^N$ be a vector indicating which Supreme Court case has been cited at time $t$, where $c_i(t)=1, i \in \{1, \dots , N\}$ indicates that the $i$th case has been cited at time $t$ and $c_i(t)=0$ indicates that the $i$th case has not been cited at time $t$. Furthermore, let
$$\mathcal{C}_t(N)=\{c(t)\in \{0,1\}^N: c_i(t)\in \{0,1\} \}$$ 
be the set of all possible citation combinations at time $t$. Note that the cardinality of $\mathcal{C}_t(N)$ increases exponentially for every newly added case, which results in $2^N$ elements.\\
The probability function of the ERCM is defined as
\begin{equation}
P_{\theta}(c(t)~|~c(t-))=\cfrac{\exp\big(\theta^T\cdot h\big(c(t)~|~c(t-)\big)\big)}{\sum_{c(t)^*\in \mathcal{C}}\exp\big(\theta^T\cdot h\big(c(t)~|~c(t-)\big)\big)}
\label{ercm}
\end{equation}
where $c_{t-}\in \{0,1\}^{N \times (t-1)}$ is a matrix that indicates which cases have been citing in each other before time $t$, $\theta \in \mathbb{R}^q$ is a q-dimensional vector of parameters,  $h: \mathcal{C}_t(N) \to \mathbb{R}^q ~, ~ \c(t) \to (h_1(c(t)), \dots , h_q(c(t)))^T$ is a q-dimensional vector of different statistics and $\kappa(\theta) := \sum_{c(t)^*\in \mathcal{C}}\exp(\theta^T\cdot h(c(t)|c(t-)))$ is a normalization constant that ensures that (\ref{ercm}) defines a probability function on $\mathcal{C}_t$.\\[0.3cm]

% Network Statistics
The generative process of a model are informed by
the decision regarding which network statistics $h(\cdot)$ are incorporated. We include the following statistics for the Supreme Court citation network:
$$h_{edges}:\mathcal{C}(N)\to \mathbb{R}~~~, ~~~c(t) \to \sum_{i=1}^Nc_i(t)$$
the number of citations made at time $t$. 
$$h_{outstar}:\mathcal{C}(N)\to \mathbb{R}~~~, ~~~c(t) \to \sum_{j<i}^Nc_i(t)\cdot c_j(t) \cdot \sqrt{\dfrac{(t-a)(t-a-b)}{t^2}}$$
the number of weighted outstars occuring at time $t$. We argue that it should me more likely to cite more recent cases than cases that have been decided further in the past. For the weight 
$$w(a,b):= \sqrt{\dfrac{(t-a)(t-a-b)}{t^2}}$$
we define $a$ and $b$ as the elapsed time since case $i$ and $j$ have been introduced to the network.
$$h_{triangle}:\mathcal{C}(N)\to \mathbb{R}~~~, ~~~c(t) \to \sum_{j<i}^Nc_i(t)\cdot c_j(t) \cdot
c_j(t_{-i}) \cdot w(a,b)$$
where $c_j(t_{-i})$ indicates whether case $j$ was cited at the time case $i$ was introduced into the network. Just as for the outstar statistic, we include a weighting factor to favor more recent cases. \\[0.3cm]

@@

\subsection*{Maximum Likelihood Estimator}
The more rigorous technique is to estimate the parameters directly with the log-likelihood function derived from (\ref{ercm}), which has the following form:
%
\begin{equation}
\text{loglik}(\theta)=\theta^T \cdot h(c(t)| c(t-))-\log(\kappa(\theta))
\label{loglik}
\end{equation}
%
The problem resulting from estimating the parameters with (\ref{loglik}) is that the term
%
$$\kappa(\theta):= \sum_{c(t)^* \in \mathcal{C}(N)} \exp(\theta^T \cdot h(c(t)^*|c(t-)))$$ 
%
which sums up the weighted statistics of all possible binar vectors of length $N$, has to be evaluated.

\subsection*{MCMC}

>>>>

 \section*{The Exponential Random Configuration Model}
Let $c(t)\in \{0,1\}^N$ be a vector indicating which Supreme Court case has been cited at time $t$, where $c_i(t)=1, i \in \{1, \dots , N\}$ indicates that the $i$th case has been cited at time $t$ and $c_i(t)=0$ indicates that the $i$th case has not been cited at time $t$. Furthermore, let
$$\mathcal{C}_t(N)=\{c(t)\in \{0,1\}^N: c_i(t)\in \{0,1\} \}$$ 
be the set of all possible citation combinations at time $t$. Note that the cardinality of $\mathcal{C}_t(N)$ increases exponentially for every newly added case, which results in $2^N$ elements.\\
The probability function of the ERCM is defined as
\begin{equation}
P_{\theta}(c(t)~|~c(t-))=\cfrac{\exp\big(\theta^T\cdot h\big(c(t)~|~c(t-)\big)\big)}{\sum_{c(t)^*\in \mathcal{C}}\exp\big(\theta^T\cdot h\big(c(t)~|~c(t-)\big)\big)}
\label{ercm}
\end{equation}
where $c(t-)\in \{0,1\}^{N \times (t-1)}$ is a matrix that indicates which cases have been citing in each other before time $t$, $\theta \in \mathbb{R}^q$ is a q-dimensional vector of parameters,  $h: \mathcal{C}_t(N) \to \mathbb{R}^q ~, ~ \c(t) \to (h_1(c(t)), \dots , h_q(c(t)))^T$ is a q-dimensional vector of different statistics and $\kappa(\theta) := \sum_{c(t)^*\in \mathcal{C}}\exp(\theta^T\cdot h(c(t)|c(t-)))$ is a normalization constant that ensures that (\ref{ercm}) defines a probability function on $\mathcal{C}_t$.\\[0.3cm]

% Network Statistics
The generative process of a model are informed by
the decision regarding which network statistics $h(\cdot)$ are incorporated. We include the following statistics for the Supreme Court citation network:
$$h_{edges}:\mathcal{C}(N)\to \mathbb{R}~~~, ~~~c(t) \to \sum_{i=1}^Nc_i(t)$$
the number of citations made at time $t$. 
$$h_{outstar}:\mathcal{C}(N)\to \mathbb{R}~~~, ~~~c(t) \to \sum_{j<i}^Nc_i(t)\cdot c_j(t) \cdot \sqrt{\dfrac{(t-a)(t-a-b)}{t^2}}$$
the number of weighted outstars occuring at time $t$. We argue that it should be more likely to cite more recent cases than cases that have been decided further in the past. For the weight 
$$w(a,b):= \sqrt{\dfrac{(t-a)(t-a-b)}{t^2}}$$
we define $a$ and $b$ as the elapsed time since case $i$ and $j$ have been introduced to the network.
$$h_{triangle}:\mathcal{C}(N)\to \mathbb{R}~~~, ~~~c(t) \to \sum_{j<i}^Nc_i(t)\cdot c_j(t) \cdot
c_j(t_{-i}) \cdot w(a,b)$$
where $c_j(t_{-i})$ indicates whether case $j$ was cited at the time case $i$ was introduced into the network. Just as for the outstar statistic, we include a weighting factor to favor more recent cases. \\[0.3cm]

@@

\subsection*{Maximum Likelihood Estimator}
The more rigorous technique is to estimate the parameters directly with the log-likelihood function derived from (\ref{ercm}), which has the following form:
%
\begin{equation}
\text{loglik}(\theta)=\theta^T \cdot h(c(t)| c(t-))-\log(\kappa(\theta))
\label{loglik}
\end{equation}
%
The problem resulting from estimating the parameters with (\ref{loglik}) is that the term
%
$$\kappa(\theta):= \sum_{c(t)^* \in \mathcal{C}(N)} \exp(\theta^T \cdot h(c(t)^*|c(t-)))$$ 
%
which sums up the weighted statistics of all possible binar vectors of length $N$, has to be evaluated. However, the cardinality of $\mathcal{C}(N)$ (\#$(\mathcal{C})=2^N$) is incredibly large and a direkt calculation of this sum is for already small $N$ not feasible. \\[0.3cm]
An solution for this limitation is based on the following consideration: Fix a vector of parameters $\theta_0 \in \Theta$ from the underlying parameter range $\Theta$ and compute for $\theta \in \Theta$ the expected value
%
\begin{eqnarray*}
\mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(C(t))\right) \right]&=& 
\sum_{c(t) \in \mathcal{C}(N)}\exp\left((\theta - \theta_0)^T \cdot \Gamma(c(t))\right)\cdot \mathbb{P}_{\theta_0}(C(t)=c(t))\\
&=& \sum_{c(t) \in \mathcal{C}(N)}\exp\left((\theta - \theta_0)^T \cdot \Gamma(c(t))\right)\cdot 
\frac{\exp(\theta_0^T \cdot \Gamma(c(t)))}{\kappa(\theta_0)}\\
&=& \frac{1}{\kappa(\theta_0)} \sum_{c(t) \in \mathcal{C}(N)}\exp\left(\theta^T \cdot \Gamma(c(t))\right)\\
&=&\frac{\kappa(\theta)}{\kappa(\theta_0)}
\end{eqnarray*}
%
This equation offers the following possibility: If one draws $L$ random vectors $c^{(1)}(t), \dots ,c^{(L)}(t)$ out of a distribution $\mathbb{P}_{\theta_0}$ appropriately, one gets with the law of big numbers and a big enough sample $L$ the following relation:
%
\begin{equation}
\frac{1}{L}\cdot \sum_{i=1}^{L}  \exp\left((\theta - \theta_0)^T \cdot \Gamma(c^{(i)}(t))\right)
~~\approx~~ \mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(C(t))\right) \right] = \frac{\kappa(\theta)}{\kappa(\theta_0)}
\label{konver}
\end{equation}
%
This approximate can then be used to approximate the log likelihood function.\\[0.4cm]
Next, we will discuss how a sufficient number of suitable drawings $c^{(1)}(t), \dots ,c^{(L)}(t)$ can be sampled from the distribution $\mathbb{P}_{\theta_0}$. \\
For this purpose, the Markov Chain Monte Carlo (MCMC) methods can be used.
\subsection*{Metropolis Hastings for the ERCM}\label{networksimulation}
To be able to compute the approximate likelihood function one needs a sufficiently large number of random vectors from the distribution $\mathbb{P}_{\theta_0}$. Snijders \cite{Snijders.2002b} introduces an approach to sample random networks for the ERGM framework by using \textit{MCMC methods}. We adapt this approach for sampling appropriate binary vectors for the ERCM. \\[0.4cm]
\textit{Metropolis-Hastings Algorithm}\\\\
Choose any vector $c^{(0)}(t) \in \mathcal{C}(N)$ to start with (e.g., the observed vector). For $k \in \{0,...,L-1\}$ recursively proceed as follows:\\
\begin{enumerate}
\item Randomly choose a number $i\in \{1,\dots, N\}$
\item Compute, using the equation (\ref{Logit}) the value
$$\pi := \dfrac{\mathbb{P}_{\theta}(C_{i}(t) \neq c_{i}^{(k)}(t) ~| ~C_i(t)^c=c_i(t)^c)}{\mathbb{P}_{\theta}(C_{i}(t) = c_{i}^{(k)}(t) ~| ~C_i(t)^c=c_i(t)^c)}$$
\item Define $\delta:= \min\{1, \pi\}$ and draw a random number $Z$ from Bin$(1, \delta)$. If
\begin{itemize}
\item $Z=0$, let $c^{(k+1)}(t) := c^{(k)}(t)$ 
\item $Z=1$, define $c^{(k+1)}(t)$ as
$$c_{p}^{(k+1)}(t)=\begin{cases}
1-c_{p}^{(k)}(t)& \text{if}~ p=i \\
c_{p}^{(k)}(t) &\text{if}~ p \neq i 
\end{cases}$$
\end{itemize}
\item Start at step 1 with $c^{(k+1)}(t)$.
\end{enumerate}
The first $B \ll L$ vectors are discarded as Burn-In.
